---
title: "Using EpiStainDynamics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using-EpiStrainDynamics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, results='hide', message=FALSE}
library(EpiStrainDynamics)

# Set some stan settings
rstan::rstan_options(auto_write = TRUE)
options(mc.cores = 1)
```

EpiStrainDynamics extends an existing statistical modelling framework capable of inferring the trends of up to two pathogens. The modeling framework has been extended here to handle any number of pathogens; fit to time series data of counts (eg, daily number of cases); incorporate influenza testing data in which the subtype for influenza A samples may be undetermined; account for day-of-the-week effects in daily data; include options for fitting penalized splines or random walks; support additional (optional) correlation structures in the parameters describing the smoothness of the penalized splines (or random walks); and account for additional (optional) sources of noise in the observation process. 

# Step 1: Construct model

These modelling specifications are specified using the `construct_model` function. The correct stan model is then applied based on the specifications provided. `construct_model()` takes three arguments: `method`, `pathogen_structure`, and `dow_effect`. We will break these down one by one. 

## Method

EpiStrainDynamics has pre-compiled stan models that fit either with bayesian penalised splines or random walks. These are specified using the `method` argument of `construct_model()` as functions, either with `random_walk()` or `p_spline()`. The penalised spline model has two further options to specify: `spline_degree` is the polynomial degree of the individual spline segments used to construct the overall curve (must be a positive whole number) and `days_per_knot`, which is the number of days for each knot (must also be a positive whole number). 

So we may specify: 
```
mod <- construct_model(
  method = random_walk(), 
  ...
)
# OR

mod <- construct_model(
  method = p_spline(spline_degree = 3, days_per_knot = 2),  # example options
  ...
)
```

## Pathogen structure

There are three main types of pathogen structure available to model: `single()`, `multiple()`, and `subtyped()`. 

The `single()` pathogen structure is the simplest and models a single pathogen timeseries. A vector is the case timeseries is provided to the argument `case_timeseries`, a vector of date or time labels is provided to `time`, and optionally a pathogen name can be assigned with `pathogen_name`. It can be specified as follows, illustrated using data provided with the package `sarscov2`: 

```
single(
  case_timeseries = sarscov2$cases,           # timeseries of case data
  time = sarscov2$date,                       # date or time variable labels
  pathogen_name = 'SARS-COV-2'                # optional name of pathogen 
)
```

The `multiple()` and `subtyped()` pathogen structures both also have the `case_timeseries` and `time` arguments, as in the `single()` structure. 

In addition to these two arguments, instead of a single argument for `pathogen_name`, `multiple()` and `subtyped()` have one or more arguments that allow the user to define the names and data for the different components or subtypes to be modelled. For `multiple()`, these are specified as a named list with the argument `component_pathogen_timeseries`. The names in this list will be using in subsequent plotting. For `subtyped()`, which allows the user to incorporate testing data for influenza A subtypes, a vector of total unsubtyped influenza A cases is specified with `influenzaA_unsubtyped_timeseries`, a named list of the subtyped influenza A timeseries is provided to `influenzaA_subtyped_timeseries`, and further pathogens are provided in a named list to `other_pathogen_timeseries`. 

In addition to specifying the data and pathogen names, `multiple()` and `subtyped()` both allow the user to modify the correlation structures in the parameters describing the smoothness (with argument `smoothing_structure`) and account for additional sources of noise in the observation process (`observation_noise`). `smoothing_structure` is either `'shared'` (all pathogens have the same smoothness), `'independent'` (each pathogen has completely independent smoothing structure), or `'correlated'` (smoothing structure is correlated among pathogens). `observation_noise` is either `'observation_noise_only'` (only includes observation noise - the same between pathogens) or `'pathogen_specific_noise'` (includes noise in individual pathogens as well). 

Example pathogen structure specification for multiple pathogens model:
```
multiple(
   case_timeseries = sarscov2$cases,        # timeseries of case data
   time = sarscov2$date,                    # date or time variable labels
   
   component_pathogen_timeseries = list(    # named list of component pathogens
     alpha = sarscov2$alpha,
     delta = sarscov2$delta,
     omicron = sarscov2$omicron,
     other = sarscov2$other
   ),
   
   smoothing_structure = 'independent',             # correlation structures
   observation_noise = 'observation_noise_only'     # observation noise
 )
```

Example pathogen structure specification for subtyped model:
```
subtyped(
   case_timeseries = influenza$ili,         # timeseries of case data
   time = influenza$week,                   # date or time variable labels
   
   influenzaA_unsubtyped_timeseries = influenza$inf_A,  # unsubtyped influenzaA
   influenzaA_subtyped_timeseries = list(       # named list of subtyped infA
     H3N2 = influenza$inf_H3N2,
     H1N1 = influenza$inf_H1N1
   ),
   other_pathogen_timeseries = list(            # named list of other pathogens
     influenzaB = influenza$inf_B,
     other = influenza$num_spec - influenza$inf_all
   ),
   
   smoothing_structure = 'correlated',            # correlation structures
   observation_noise = 'pathogen_specific_noise'  # observation noise
 )
```

## Day of week effect

Day of week effect is specified as a logical (`TRUE` or `FALSE`) to the `dow_effect` argument. In plotting the day of week effect can be selectively removed. 

```
mod <- construct_model(
  method = random_walk(), 
  pathogen_structure = ...,
  dow_effect = TRUE
)
```

```{r, echo=FALSE}
pathogen_structure = subtyped(
    case_timeseries = influenza$ili,
    time = influenza$week,
    influenzaA_unsubtyped_timeseries = influenza$inf_A,
    influenzaA_subtyped_timeseries = list(
      H3N2 = influenza$inf_H3N2,
      H1N1 = influenza$inf_H1N1
    ),
    other_pathogen_timeseries = list(
      influenzaB = influenza$inf_B,
      other = influenza$num_spec - influenza$inf_all
    ),
    smoothing_structure = 'independent',
    observation_noise = 'observation_noise_only'
  )
mod <- construct_model(
  method = p_spline(),
  pathogen_structure = pathogen_structure,
  dow_effect = TRUE
)
```

# Step 2: fit model

The model estimates the expected value of the time series (eg, a smoothed trend in the daily number of cases accounting for noise) for each individual pathogen. Model parameterisation decisions specified when constructing the model in step 1 mean the correct stan model will be applied at this stage by simply calling `fit_model()` onto the constructed model object. 

```{r, cache = TRUE, results = 'hide'}
fit <- fit_model(
  mod,
  iter = 2000,
  warmup = 1000,
  chains = 3
)
```

# Step 3: Calculate and visualise epidemiological quantities

Calculate epidemic growth rate with `growth_rate()`:
```{r}
gr <- growth_rate(fit)
plot(gr)
```

Calculate effective reproduction number over time with `Rt()` (requiring specification of a generation interval distribution): 
```{r}
rt <- Rt(fit, gi_dist = function(x) 4*x*exp(-2*x))
plot(rt)
```

Calculate incidence with or without a day of week effect with `incidence()`:
```{r}
inc_no_dow <- incidence(fit, dow = FALSE)
plot(inc_no_dow)
```

```{r}
inc_dow <- incidence(fit, dow = TRUE)
plot(inc_dow)
```

Calculate proportions of different combinations of cases attributable to different pathogens/subtypes using `proportion()`, where the pathogens/subtypes are specified by their names in the named lists provided to `construct_model()`:
```{r}
proportion(fit, numerator_combination = c('H3N2', 'H1N1', 'influenzaB'),
           denominator_combination = c('H3N2', 'H1N1', 'influenzaB', 'other'))
```

